Games with ordinal sequences of moves

### 2016-10-11 ordinal_games/split-epimorphisms/weakenings.tex
Started Research repository

\input{../common/preamble}

\title{Games with ordinal sequences of moves}
\author{John Gowers}
\date{}

@@

\begin{titlepage} % Shamelessly stolen from the online LaTeX book.
	\centering
	{\scshape\LARGE University of Bath \par}
	\vspace{1cm}
	{\scshape\Large PhD Confirmation Report \par}
	\vspace{1.5cm}
	{\huge\bfseries Games with ordinal sequences of moves \par}
	\vspace{2cm}
	{\Large\itshape John Gowers \par}
	\vfill
	supervised by\par
	Dr.~James \textsc{Laird}

	\vfill

% Bottom of the page
	{\large \today\par}
\end{titlepage}

\addtocontents{toc}{\protect\enlargethispage{35mm}}
\tableofcontents
\thispagestyle{empty}

\setcounter{tocdepth}{2}

\section{Introduction}

\lettrine[image=true, lines=6, findent=3pt, nindent=0pt]{cleos-g.png}{ame Semantics}, or the use of games and strategies to model logical systems and programming languages, has its origins in a model of intuitionistic logic by Lorenzen \cite{lorenzengames} using games.  Towards the end of the twentieth century, several authors, starting with Blass \cite{blassgames}, realized that certain natural operations on games could be used to model the connectives in Girard's linear logic.  Major results from this enterprise include:
\begin{itemize}
  \item A fully complete categorical semantics for the multiplicative fragment of linear logic, together with the MIX rule, by Abramsky and Jagadeesan \cite{abramskyjagadeesangames}
  \item A fully complete categorical semantics for the multiplicative fragment of linear logic without the MIX rule, by Hyland and Ong \cite{HylandOngGames}
  \item A modification of Abramsky and Jagadeesan's construction, giving a fully complete model of the multiplicative-exponential fragment of linear logic, by Baillot, Danos, Ehrhard and Regnier \cite{bion}
  \item Fully complete models of the negative fragment of linear logic by others, starting with independent work by Curien \cite{curiengames} and Lamarche \cite{lamarchegames}.  
\end{itemize}

Another major early result was the use of game semantics to give a fully abstract model of the programming language PCF (This result was obtained independently, using different techniques, by Abramsky, Jagadeesan and Malacaria \cite{ajmPcf}, by Hyland and Ong \cite{hoPcf}, and by Nickau \cite{nickauPcf}).

Central to much of the current work on game semantics for linear logic is the treatment of the exponential connective $\oc$.  Several approaches have been made to model this connective in a game-theoretic way (see Section 3 of \cite{martinsthesis} for a brief survey).  We shall use the construction from \cite{ajmPcf} and \cite{hyland1997games}: if $A$ is a game, then $\oc A$ is the game formed by playing a countably infinite number of copies of $A$ --- $A_0,A_1,A_2,\dots$ --- in parallel, where only the opponent $O$ may switch games and where neither player may play in game $A_{n+1}$ until a move has been made in game $A_n$.

This definition, and in particular the restriction on the order in which games may be started, allows us to give the game $\oc A$ the structure of the \emph{cofree commutative comonoid} over $A$ \cite{LairdCofCommCom} and gives rise to a \emph{linear exponential comonad} on our category of games.  The linear exponential comonad \cite{SchalkWhatIs} and the cofree commutative comonoid functor \cite{LafontCofCommCom,MelliesCofCommCom} are both sufficient to model the exponential connective from linear logic.

The restriction that says that the copies of $A$ in $\oc A$ must be opened in order, motivates the following definition \cite{laird02}: if $A,B$ are games, then the \emph{sequoid} $A\sequoid B$ of $A$ and $B$ is the game in which $A$ and $B$ are played in parallel, only the opponent $O$ may switch games, and play must start in $A$.  It turns out that if $A$ is a game, then the operation $A\sequoid\blank$ gives rise to an endofunctor on the category $\G$ of games, and that the exponential $\oc A$ is the final coalgebra for this endofunctor.  That is to say, we have a morphism $\cmap{\alpha}{\oc A}{A\sequoid\oc A}$ such that whenever $\cmap{\sigma}{B}{A\sequoid B}$ is a morphism, there is a unique morphism $\cmap{\fcoal{\sigma}}{B}{\oc A}$ such that the following diagram commutes:
\[
  \begin{tikzcd}
    B \arrow[r, "\sigma"] \arrow[d, "\fcoal\sigma"']
      & A\sequoid B \arrow[d, "A\sequoid\fcoal\sigma"] \\
    \oc A \arrow[r, "\alpha"']
      & A\sequoid\oc A
  \end{tikzcd}
  \]
Existing proofs of this fact in a finitary games setting (e.g., in \cite{martinsthesis}) make implicit use of the fact that $\oc A$ is the limit of the sequence
\[
  I \leftarrow A \leftarrow A\sequoid A \leftarrow A\sequoid(A\sequoid A) \leftarrow \dots
  \]
Indeed, any finite-length play in $\oc A$ must involve only finitely many copies of $A$, so exists as a play in $(A\sequoid\blank)^n I$ for some $n$.

On the other hand, when we start imposing winning conditions upon infinite plays, $\oc A$ is no longer the limit of the sequence above.  Indeed, define a game $\co A$ which is exactly like $\oc A$ except that all infinite plays that contain moves from infinitely many copies of $A$ are wins for the player $P$, regardless of who has won the individual games.  This game still admits a cone over the sequence above, since any play occurring in $(A\sequoid\blank)^nI$ for some $n$ must involve only finitely many copies of $A$.  Moreover, it does not admit a morphism of cones to $\oc A$.  In fact, in the infinitary setting, $\co A$, and not $\oc A$, is the limit of the above sequence.

\[
  \begin{tikzcd}
    \co A \arrow[d] \arrow[dr] \arrow[drr] \arrow[drrr] \arrow[drrrr, phantom, yshift=0.5em, "\cdots"]
      &
        &
          &
            & \\
    I
      & A \arrow[l]
        & A\sequoid A \arrow[l]
          & A\sequoid(A\sequoid A) \arrow[l]
            & \cdots \arrow[l]
  \end{tikzcd}
  \]
Applying the functor $A\sequoid\blank$ to the cone from $\co A$ gives rise to another cone over the sequence:
\[
  \begin{tikzcd}
    %
    & A\sequoid\co A \arrow[dl] \arrow[d] \arrow[dr] \arrow[drr] \arrow[drrr, phantom, yshift=0.5em, "\cdots"]
        &
          &
            & \\
    I
      & A \arrow[l]
        & A\sequoid A \arrow[l]
          & A\sequoid(A\sequoid A) \arrow[l]
            & \cdots \arrow[l]
  \end{tikzcd}
  \]
and therefore a morphism $A\sequoid\co A\to \co A$.  We can then continue to apply the functor $A\sequoid\blank$ to build up a second layer of the sequence:
\[
  \begin{tikzcd}
    \co A \arrow[d] \arrow[dr] \arrow[drr] \arrow[drrr] \arrow[drrrr, phantom, xshift=1em, "\cdots"]
      & A\sequoid\co A \arrow[l]
        & A\sequoid(A\sequoid \co A) \arrow[l]
          & A \sequoid(A\sequoid(A\sequoid\co A)) \arrow[l]
            & \cdots \arrow[l] \\
    I
      & A \arrow[l]
        & A\sequoid A \arrow[l]
          & A\sequoid(A\sequoid A) \arrow[l]
            & \cdots \arrow[l]
  \end{tikzcd}
  \]
What is the limit of this entire diagram?  An $O$-winning infinite play $s$ in $(A\sequoid\blank)^n \co A$ may be identified with an $O$-winning play in $\oc A$ such that either the opponent $O$ wins in one of the games $A_1,\dots,A_n$ or that only finitely many copies of $A$ are opened.  It follows that the limit of the diagram \emph{is} the final coalgebra $\oc A$.

The diagrams we have drawn above are special cases of the \emph{final sequence} \cite{finalseq}: given a category $\C$ with a terminal object $1$ and enough limits (in a sense which will become clear) and an endofunctor $\cmap{\F}{\C}{\C}$, we may construct an ordinal-indexed sequence
\[
  1, \F(1), \F^2(1),\dots \F^\omega(1), \F^{\omega+1}(1),\dots,\F^\delta(1),\dots
  \]
together with commuting morphisms $\F^\gamma(1)\to\F^\delta(1)$ for $\delta\le\gamma$, by setting $\F^{\delta+1}(1) = \F(\F^\delta(1))$ and letting $\F^\mu(1)$ be the limit of the diagram formed by the $\F^\gamma(1)$ for $\gamma<\mu$ if $\mu$ is a limit-ordinal.  It is well known that if $\delta$ is an ordinal and the map
\[
  \F^{\delta+1}(1)\to\F^{\delta}(1)
  \]
is an isomorphism, then the inverse of this isomorphism exhibits $\F^{\delta}(1)$ as the final coalgebra of $\F$.  In this case, we say that the final sequence \emph{stabilizes at $\delta$}.

In the case of finitary games, then, we have
\[
  \oc A\cong(A\sequoid\blank)^\omega(I)
  \]
and in the infinitary case, we have
\[
  \oc A\cong(A\sequoid\blank)^{\omega2}(I)
  \]
The eventual goal of this research is to study the behaviour of the final sequence of the functor $A\sequoid\blank$ in categories of games.  Since the sequence stabilizes so early on for existing categories of games, this question is not very interesting in that setting.  We want to modify our definitions of games in such a way that this final sequence will take much longer to stabilize, and it turns out that one way to do that is to allow plays of transfinite length.

Games with ordinal sequences of moves are not a new idea.  They have, for example, been studied in \cite{longgamesbook} from the point of view of determinacy.  Even in game semantics, as we shall see, we may identify games in which we impose a winning condition on the infinite plays with games in which the players are allowed to make plays of length $\omega$ -- we shall call these $(\omega+1)$-games, in contrast to the finitary games, which we shall call $\omega$-games.  Allowing plays whose lengths are longer transfinite ordinals will give rise to a richer variety of plays, which will be reflected in the final sequence.

One important aspect of the shift from finite to transfinite plays is that the idea of a \emph{play} or \emph{position} takes on a new importance, in contrast to the finitary case, where we can define everything in terms of \emph{moves}.  Normally, a move is designated as either a $P$-move or an $O$-move; we will extend this so that every position is either a $P$-position or an $O$-position.  A position $s$ will be an ordinal-indexed sequence of moves; if there is a last move $a$, then $s$ is a $P$-position or an $O$-position according as $a$ is a $P$-move or an $O$-move.  If the length of $s$ is a limit ordinal, however, then its sign cannot be determined from the signs of its constituent moves.  This is useful if we want to model games with a winning condition as $(\omega+1)$-games: a play of length $\omega$ in a game corresponds to an infinite play in that game, and its designation as a $P$-position or $O$-position is essentially the same thing as saying whether it is a $P$-winning or $O$-winning play.

This approach is similar to that of Blass \cite{blassgames}, and indeed our treatment of $P$-positions and $O$-positions under the connectives will follow his approach.  For example, a position $s$ in a tensor product $A\tensor B$ is a $P$-position if and only if its restrictions to $A$ and $B$ are both $P$-positions.

To avoid certain mathematical difficulties, we shall follow Curien, Lamarche and others by requiring that our games be \emph{negative} -- that is, they must always begin with an $O$-move.  In fact, we shall go further than that: under the usual alternating condition, saying that $O$ always starts is the same thing as requiring that the empty position $\emptyplay$ is a $P$-position.  In order to avoid problems constructing our category, we shall need to require that \emph{all} positions whose lengths are limit ordinals be $P$-positions.  We shall call games satisfying this property \emph{completely negative}.

This condition on the sign of games allows us to construct a well-behaved category of games, but it has one serious problem that occurs as soon as we consider games with transfinite-length plays: if $A,B$ are completely negative games, then the linear implication $A\implies B$ is not necessarily completely negative.

We want $A\implies B$ to be the internal hom in our closed monoidal category, so we need to find a way around this.  Our solution is to construct, for every game $A$, a `minimal extension' $A^{\cn}$ of $A$ that is completely negative.  We prove various results about the relationship $A^{\cn}$ and $A$ that make this precise and then use them to show that the game $A\impliescn B\coloneqq (A\implies B)^{\cn}$ does give rise to a closed monoidal structure in our category.  We also give a tentative way of constructing a category whose objects are \emph{all} games, not just the completely negative ones, and use our results to show that this category is in fact equivalent to our existing category of completely negative games.

Having constructed a symmetric monoidal closed category of games, we introduce the sequoid operator and the exponential, and study the final sequence.  We shall characterize the exponential $\oc A$ as a final coalgebra for the functor $A\sequoid\blank$ and use this characterization to prove that $\oc A$ admits the structure of a cofree commutative comonoid on $A$.  Then we turn to the final sequence.  This section will involve some material that is more ordinal-theoretic than game-theoretic; we assign an ordinal \emph{rank} to a (transfinite) sequence of natural numbers and then show that this idea of rank can be used to characterize each term of the final sequence for a functor of the form $A\sequoid\blank$.  We will show that the reason that the final sequences converge so quickly for finitary games and win-games is that a sequence of length $\omega$ necessarily has low rank: once we allow longer plays, we can get sequences with higher ranks and the final sequence will carry on for longer.

\subsection{Original contribution}

Games with ordinal sequences of enablings have been studied by Laird from a Game Semantics point of view \cite{LairdOrdinalGames}, but to my knowledge this is the first transfinite games category built in the tradition of Abramsky and Jagadeesan.  Passing to transfinite games causes problems when we try to show that our category is monoidal closed, and we have to come up with new techniques in order to get a monoidal closed category (see Sections \ref{TransSMCCSex} to \ref{ImpliesCnSex}).

The characterization of the final coalgebra for the functor $A\sequoid\blank$ in a transfinite games category is new, though the behaviour in finitary game categories and win-game categories is well known.  Our definition of the rank of a sequence of natural numbers is heavily motivated by the game semantics, so it is likely that this is new too (though similar constructions of ordinal ranks of objects are common in mathematics).

Otherwise, most of the material in this report is a straightforward translation of the existing game semantics of Abramsky, Jagadeesan, Hyland, Ong, Curien, Lamarche and so on into the transfinite setting.

\subsection{Conventions}

I shall denote the composition of morphisms $A\xrightarrow{\sigma}B\xrightarrow{\tau}C$ by $\comp\tau\sigma$, the disjoint union of two sets $X,Y$ by $X\cprd Y$ and the corresponding copairing of two functions $\map{f}{X}{Z},\map{g}{Y}{Z}$ by $\cmap{f\cprd g}{X\cprd Y}{Z}$.

Following the $\forall$belard/$\exists$loise convention that is sometimes used for Opponent/Player, I shall refer to the opponent $O$ as `he' and the player $P$ as `she' throughout.

\subsection{Acknowledgements}

It goes without saying that I owe my biggest debt of gratitude for the material in this report to my supervisor, Jim Laird.  Dr Laird introduced me to Game Semantics a year ago and suggested my initial direction of research in the subject, and his insights and instinct for the subject have been invaluable in the development of the material contained in this report, and in the preparation of the report itself -- particularly when things were going wrong.  Dr Laird gave me some extremely useful feedback on an earlier draft of this report and supplied some of the details for the material in section \ref{SeqExpIsCofCommCom}.

I am indebted to all the authors listed in the Bibliography, but I single out the seminal papers of Blass \cite{blassgames} and Abramsky-Jagadeesan \cite{abramskyjagadeesangames} for their development of the subject and their clear presentation of their full completeness results, the set of notes on Game Semantics by Martin Hyland \cite{hyland1997games} and the excellent online notes by Andrea Schalk \cite{Schalk2001GsNotes,SchalkWhatIs}.  I also single out the PhD thesis of Martin Churchill \cite{martinsthesis}, prepared under Dr Laird's supervision.  The work in this report can be seen as a partial sequel to some of Dr Churchill's work, and the clear presentation in his thesis has greatly helped in this task.

I would also like to extend my thanks to Paul Blain Levy for his work organizing the GaLoP 2016 conference and the Midlands Graduate School 2016, both of which I attended this year.  I presented some of this material at GaLoP and the questions and comment I received there have shaped the course of this work.  Thanks also to Alexander Kurz for his lectures on coalgebra at the Midlands Graduate School.

Thanks go to all my friends and family members who have patiently supported me while I embarked on this research.  Thank you to CN for designing the splendid letter `G' (complete with $\sequoid$ symbols!) that features at the beginning of this introduction.

\section{Our starting category of games}

Before studying games with transfinite sequences of moves, we shall illustrate some of the choices we have made by defining a category of games with finite sequences of moves.  We have chosen these definitions because they extend particularly well to the transfinite case.

\subsection{Games and strategies}

We shall use the notation introduced in \cite{abramskyjagadeesangames} to describe games.  All our games $A$ will have, at their heart, the following three pieces of information:
\begin{itemize}
  \item A set $M_A$ of possible moves
  \item A function $\cmap{\lambda_A}{M_A}{\OP}$ assigning to each move the player who is allowed to make that move
  \item A prefix-closed set $P_A\subset M_A^*$ of finite sequences of moves.
\end{itemize}
We shall normally insist on an \emph{alternating condition} on $P_A$:
\begin{description}
  \item[Alternating condition] If $a,b\in M_A$ are moves and $s\in M_A^*$ is a sequence of moves such that $sa, sab\in P_A$, then $\lambda_A(a)=\neg\lambda_A(b)$.
\end{description}

As in \cite{abramskyjagadeesangames}, we identify a \emph{strategy} for a game $A$ with the set of sequences of moves that can occur when player $P$ is playing according to that strategy so that a typical definition of a (partial) strategy might be a set $\sigma\subset P_A$ such that (for all $s\in M_A^*, a,b\in M_A$):
\begin{itemize}
  \item $\emptyplay\in\sigma$ (ensures that $\sigma$ is non-empty)
  \item If $sa\in\sigma$, $\lambda_A(a)=P$ and $sab\in P_A$ then $sab\in\sigma$ ($\sigma$ contains all legal replies by player $O$)
  \item If $s,sa,sb\in\sigma$ and $\lambda_A(a)=P$ then $a=b$ ($\sigma$ contains at most one legal reply by player $P$)
\end{itemize}

We can impose additional constraints on $\sigma$ that will ensure that $\sigma$ is total, strict, history free and so on.  The definition given immediately above is not the only definition of a strategy found in the literature, however.  For example, the games described in \cite{abramskyjagadeesangames} have the curious property that the set $P_A$ may contain plays that cannot actually occur when $A$ is being played; in particular, all plays must start with a move by player $O$, but the set $P_A$ may contain positions that start with a $P$-move.  These plays do not affect the strategies for $A$, but they might come into play if we perform operations on $A$ such as forming the negation $\neg A$ or the implication $A\implies B$.

This behaviour is made implicit in Abramsky and Jagadeesan's definitions, which do not impose any conditions upon the set $P_A$ beyond the basic alternation condition given above, but which mandate that any play occurring \emph{in a strategy} must begin with an $O$-move.  For the sake of clarity, we adopt a different, but completely equivalent, approach.  For a game $A$, we define a set $L_A$, regarded as the set of \emph{legal plays} occurring in $P_A$.  In some games models, such as that found in \cite{blassgames}, $L_A$ may be defined to be the whole of $P_A$, while in \cite{abramskyjagadeesangames} it is defined to be that subset of $P_A$ consisting of plays that begin with an $O$-move.

The point of specifying $L_A$ separately is that it allows us to unify the definition of a \emph{strategy}, while making clearer the behaviour observed above, whereby certain plays in $P_A$ may not occur `in normal play'; this behaviour was previously only implicit in the definition of a strategy.  Our unified definition then becomes:

\begin{definition}
  If $A=(M_A,\lambda_A,P_A)$ is a game and $L_A$ is its associated set of legal plays (in a particular games model) then a (partial) \emph{strategy} for $A$ is a subset $\sigma\subset L_A$ such that for all $s\in L_A$ and all $a, b\in M_A$:
  \begin{itemize}
    \item $\emptyplay\in\sigma$
    \item If $s\in\sigma$ and $a$ is an $O$-move, and if $sa\in L_A$, then $sa\in\sigma$
    \item If $s\in\sigma$ and $a,b$ are $P$-moves, and if $sa,sb\in\sigma$, then $a=b$
  \end{itemize}
\end{definition}

\subsection{Positive and negative games, ownership of plays and connectives}

Abramsky-Jagadeesan games, as described in \cite{abramskyjagadeesangames}, may admit both plays that start with a $P$-move and plays that start with an $O$-move.  Other games models, such as those found in \cite{blassgames} and \cite{curiengames}, are more restrictive.  The games in \cite{curiengames} only contain plays starting with an $O$-move.  The plays in \cite{blassgames} may start with either a $P$-move or an $O$-move, but a play starting with a $P$-move and a play starting with an $O$-move may not occur in the same game.

\begin{definition}
  We say that a game $A=(M_A,\lambda_A,P_A)$ is \emph{positive} if every play in $P_A$ begins with a $P$-move.
  We say that $A$ is \emph{negative} if every play in $P_A$ begins with an $O$-move.
\end{definition}

So the Curien model found in \cite{curiengames} admits only negative games, the Blass model in \cite{blassgames} admits positive and negative games, while the Abramsky-Jagadeesan model found in \cite{abramskyjagadeesangames} admits not only positive and negative games, but also games that are neither negative nor positive.  We shall now examine the reasons for and drawbacks of each of these choices.

The earliest games model, found in \cite{conwaygames}, did not include a definition of which player is to move at a given position; rather, games are defined recursively as pairs of games $\{L|R\}$, where $L$ represents the positions that the left player may move into, while $R$ represents the positions that the right player may move into.  Blass's definition departs completely from this tradition; now, at every position $s$ only one of the two players is allowed to move; extending this logic on to the empty position $\emptyplay$, it follows that all games are either positive or negative.  This property means that we may freely define $L_A=P_A$, since there is never any question about whose turn it is to play.  By contrast, if we were to define $L_A=P_A$ for Abramsky-Jagadeesan games, then a strategy might end up containing two branches, one of plays beginning with an $O$-move and one of plays beginning with a $P$-move, which is undesirable.  The alternative definition of $L_A$ avoids this problem.

In the case of a Blass game $A$, we may define a function $\zeta_A\colon P_A\to\OP$ that says which player owns each play; the idea is that if we are in position $s$, then the next player to move is given by $\neg\zeta_A(s)$; i.e., the opposing player to the player who has just made the move.  One might want to define $\zeta_A$ by setting $\zeta_A(sa)=\lambda_A(a)$, so that ownership of a play is decided by who has made the last move in the play, but this definition does not extend in an obvious way to the empty position $\emptyplay$ (and, as we shall see in the next chapter, it does not extend to plays over limit ordinals).  In this case, $\zeta_A(\emptyplay)$ is part of the game's data, and it determines whether the game is positive or negative: if $\zeta_A(\emptyplay) = P$ then all plays must start with an $O$-move, and the game is negative -- and vice versa.

An important question then arises: how should we extend the function $\zeta_A$ to games formed from connectives?  The solution adopted by Blass is to use binary conjunctions to deduce the ownership of a play from the ownership of the restrictions of that play to the two component games.  In the case of the tensor product $A\tensor B$ of two games $A$ and $B$, we define $\cmap{\zeta_{A\tensor B}}{P_{A\tensor B}}{\OP}$ by setting
\[
  \zeta_{A\tensor B}(s) = (\zeta_A(s\vert_A) \wedge \zeta_B(s\vert_B))
  \]
where $\cmap{\wedge}{\OP\times \OP}{\OP}$ is as in Figure \ref{truthtables}.

\begin{figure}[h]
  \begin{center}
    $\begin{array}{cc|c}
      a & b & a \wedge b \\
      \hline
      O & O & O \\
      O & P & O \\
      P & O & O \\
      P & P & P
    \end{array}$
    \quad
    $\begin{array}{cc|c}
      a & b & a \vee b \\
      \hline
      O & O & O \\
      O & P & P \\
      P & O & P \\
      P & P & P
    \end{array}$
    \quad
    $\begin{array}{cc|c}
      a & b & a \Rightarrow b \\
      \hline
      O & O & P \\
      O & P & P \\
      P & O & O \\
      P & P & P
    \end{array}$
    \caption{Truth tables for binary conjunctions on $\OP$}
    \label{truthtables}
  \end{center}
\end{figure}

Similarly, we may extend $\zeta$ to the implication $A\implies B$ and the par $A\parr B$ by setting
\begin{align*}
  \zeta_{A\implies B}(s)=(\zeta_A(s\vert_A)\Rightarrow \zeta_B(s\vert_B))\\
  \zeta_{A\parr B}(s) = (\zeta_A(s\vert_A) \vee \zeta_B(s\vert_B))
\end{align*}

Note that if we use these definitions then the owner $\zeta_C(sa)$ of a play $sa$ might not correspond to the player $\lambda_C(a)$ who played the last move $a$.  For example, let $A,B$ be two positive games and form their tensor product $A\tensor B$.  Then we have
\[
  \zeta_{A\tensor B}(\emptyplay) = (\zeta_A(\emptyplay) \wedge \zeta_B(\emptyplay)) = O \wedge O = O
  \]
and so $A\tensor B$ is a positive game.  Player $P$ plays an opening move in one of the two games - let us say she plays the move $a$ in the game $A$.  But then we have
\[
  \zeta_{A\tensor B}(a) = (\zeta_A(a) \wedge \zeta_B(\emptyplay)) = P \wedge O = O
  \]
In other words, it is still player $P$'s turn to play!  Blass embrace this possibility and allows player $P$ to make these two moves.  In his paper, he introduces the notions of \emph{strict} and \emph{relaxed} games, where the strict games are the objects of study but the relaxed games are often used since they allow more manipulations.  In this case, the game $A\tensor B$ is defined as a relaxed game that might not satisfy the alternating condition; in the process of converting it into a strict game, these two opening moves by player $P$ are combined into a single move.

This `double move' can only occur at the start of the game, and Blass treats it as a special case in his proofs.  Perhaps unsurprisingly, this inconsistency causes major problems if we try to compose strategies.  We do not get an associative composition of strategies for $A\implies B$ with strategies for $B\implies C$ and so we do not get a categorical semantics.  An example of the failure of associativity in Blass's games model is given towards the end of \cite{abramskyjagadeesangames}.

>>>>

\usepackage[utf8]{inputenc}

\usepackage{graphicx} % support the \includegraphics command and options

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{mathtools} % for the all important \coloneqq symbol
\usepackage{url} % for \url
\usepackage{IEEEtrantools} % for \IEEEeqnarray

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%% Font things %%
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{cmll} % Linear logic symbols!

%% Lists %%
\usepackage{enumerate}

%% Graphics %%
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{patterns}
\usetikzlibrary{calc}

%% Theorems! %%
\usepackage{amsthm}
\theoremstyle{plain} % Theorems, lemmas, propositions etc.
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{construction}[theorem]{Construction}
\theoremstyle{definition} % Definitions etc.  Remarks too, because I don't like the way the 'remark' style looks.
\newtheorem{definition}[theorem]{Definition}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{question}[theorem]{Question}

%% Exercises and answers %%
\usepackage{answers}

\newtheoremstyle{exercisestyle}% name
  {6pt}   % ABOVESPACE
  {6pt}   % BELOWSPACE
  {\itshape}  % BODYFONT
  {0pt}       % INDENT (empty value is the same as 0pt)
  {\bfseries} % HEADFONT
  {.}         % HEADPUNCT
  {3pt} % HEADSPACE
  {}          % CUSTOM-HEAD-SPEC

\theoremstyle{exercisestyle}
\newtheorem{exercise}{Exercise}
\newtheorem{answerthm}{Exercise}

\Newassociation{answer}{answerthm}{answers}
\newcommand{\answerthmparams}{}

%% Changes to enumerate things so they look better %%

\makeatletter
\def\enumfix{%
\if@inlabel
 \noindent \par\nobreak\vskip-\topsep\hrule\@height\z@
\fi}

\let\olditemize\itemize
\def\itemize{\enumfix\olditemize}
\let\oldenumerate\enumerate
\def\enumerate{\enumfix\oldenumerate}

%% Random crap %%
\usepackage{xifthen}

\makeatletter
\def\thm@space@setup{%
  \thm@preskip=\parskip \thm@postskip=0pt
}
\makeatother

\makeatletter
\newcommand*{\relrelbarsep}{.386ex}
\newcommand*{\relrelbar}{%
  \mathrel{%
    \mathpalette\@relrelbar\relrelbarsep
  }%
}
\newcommand*{\@relrelbar}[2]{%
  \raise#2\hbox to 0pt{$\m@th#1\relbar$\hss}%
  \lower#2\hbox{$\m@th#1\relbar$}%
}
\providecommand*{\rightrightarrowsfill@}{%
  \arrowfill@\relrelbar\relrelbar\rightrightarrows
}
\providecommand*{\leftleftarrowsfill@}{%
  \arrowfill@\leftleftarrows\relrelbar\relrelbar
}
\providecommand*{\xrightrightarrows}[2][]{%
  \ext@arrow 0359\rightrightarrowsfill@{#1}{#2}%
}
\providecommand*{\xleftleftarrows}[2][]{%
  \ext@arrow 3095\leftleftarrowsfill@{#1}{#2}%
}
\makeatother

\newcommand{\catname}[1]{{\normalfont\textbf{#1}}}
\newcommand{\Rings}{\catname{CRing}}
\newcommand{\CAT}{\catname{CAT}}
\newcommand{\Top}{\catname{Top}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Cont}{\catname{Cont}}
\newcommand{\Sch}{\catname{Sch}}
\newcommand{\Rel}{\catname{Rel}}
\newcommand{\Mod}[1][]{\ifthenelse{\isempty{#1}}{\catname{Mod}}{#1\catname{mod}}}
\DeclareMathOperator{\sh}{Sh}
\newcommand{\Sh}[1][]{\ifthenelse{\isempty{#1}}{\sh}{\sh(#1)}}
\newcommand{\map}[3]{#2\xrightarrow{#1} #3}
\newcommand*\from{\colon}
\newcommand{\cmap}[3]{#1\from{}#2\to{}#3}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\DeclareRobustCommand{\vmap}[3] {\begin{tikzcd} #2 \arrow[d, "#1"] \\ #3 \end{tikzcd}}
\newcommand{\partref}[1]{(\ref{#1})}
\newcommand{\intgrpd}[4] {#1 \xrightrightarrows[#3]{#4} #2}
\DeclareRobustCommand{\bigintgrpd}[4] {\begin{tikzcd}[ampersand replacement=\&] #1 \arrow[r, shift left=0.5ex, "#3"] \arrow[r, shift right=0.5ex, "#4"'] \& #2 \end{tikzcd}}

\usepackage{xspace}

\newcommand{\etale}{\'{e}tale\xspace}
\newcommand{\Etale}{\'{E}tale\xspace}

\def \inv {^{-1}}

\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\op}{op}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\pre}{{pre}}
\DeclareMathOperator{\et}{{\acute{e}t}}

\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Spec}{Spec}

\DeclareMathOperator{\ol}{ol}

\def\presuper#1#2%
  {\mathop{}%
   \mathopen{\vphantom{#2}}^{#1}%
   \kern-\scriptspace%
   #2} \def\presub#1#2%
  {\mathop{}%
   \mathopen{\vphantom{#2}}_{#1}%
   \kern-\scriptspace%
   #2}

%% Our things %%

\newcommand{\neggame}[1]{\presuper{\perp}{#1}}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\newcommand{\varsequoid}{\vartriangleleft}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#1 \circ #2}
\newcommand{\cprd}{\sqcup}
\newcommand{\G}{\mathcal G}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\esuchthat}{\;.\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\F}{\mathcal F}
\renewcommand{\L}{\mathcal L}
\DeclareMathOperator{\wk}{wk}
\DeclareMathOperator{\weak}{weak}
\DeclareMathOperator{\str}{str}
\newcommand{\s}{\mathfrak{s}}
\renewcommand{\t}{\mathfrak{t}}
\newcommand{\grel}[1]{\underline{#1}}
\newcommand{\prefix}{\sqsubseteq}

%% Isomorphism arrows on commutative diagrams %%
\tikzset{Isom/.style={every to/.append style={edge node={node [sloped, above, allow upside down, auto=false]{$\cong$}}}},
         Isom'/.style={every to/.append style={edge node={node [sloped, above, allow upside down, auto=false, rotate=180]{$\cong$}}}},
         Sim/.style={every to/.append style={edge node={node [sloped, above, allow upside down, auto=false]{$\sim$}}}},
         Sim'/.style={every to/.append style={edge node={node [sloped, above, allow upside down, auto=false, rotate=180]{$\sim$}}}}}

%%% END Article customizations

@@

\section{The Hyland-Schalk functor $\G\to\Rel$}
\label{hylandschalke}

Let $\G$ denote an appropriate category of games and strategies, and let $\Rel$ denote the category of sets and relations.  It is known (Hyland, Schalk, 1996) that the following define a functor $\cmap{\F}{\G}{\Rel}$:

\begin{IEEEeqnarray*}{CCC}
  \F(A) & = & P_A \\
  \F\left(\map{\sigma}{A}{B}\right) & = & \{(\s\vert_A, \s\vert_B)\suchthat \s\in\sigma\textrm{ is a $P$-play}\}
\end{IEEEeqnarray*}

It can be shown that this functor $\F$ is faithful (see Calderon and McCusker, 2010 for a proof).  This surprising fact will be very useful for us: if we want to show that a particular diagram commutes in $\G$, we can normally do so more easily by applying the functor $\F$ to the diagram and showing that the resulting diagram commutes in $\Rel$.

\begin{notation}
  We shall simplify our notation by writing $P_A$ rather than $\F(A)$ and writing $\grel{\sigma}$ rather than $\F(\sigma)$.
\end{notation}
